---
title: "Introduction to practical disease modeling: Exercise 2"
author: ""
output: 
   html_document:
      toc: true
      toc_float: true
params:
  solutions: TRUE
  latex: FALSE
header-includes:
  \usepackage{comment}
---


```{r setup, include=FALSE}
## This code along with the makefile allows us to show solutions conditional on file type
set.seed(2022-07-22)
solution_start <- function(){
	rv <- ""
	if(!params$solutions & !solution_current){
		if(params$latex) rv <- "\\begin{comment}" else rv <- "<!--"
		knitr::opts_chunk$set(eval = FALSE)
		solution_current <<- TRUE
	}
	return(rv)
}
solution_end <- function(){
	rv <- ""
	if(!params$solutions & solution_current){
		if(params$latex) rv <- "\\end{comment}" else rv <- "-->"
		knitr::opts_chunk$set(eval = TRUE)
		solution_current <<- FALSE
	}
	return(rv)
}
solution_current <- FALSE
```

Exercises with infection dynamics

***

# 1. Introducing herd dynamics


We use the same herd model as introduced earlier. This time we simulate a disease, that infects the cows at random. The disease is chronic, meaning that all susceptible cows can be infected, and will stay there. Thus, it is an *SI* model.


```{r}
set.seed(250)

n_cows <- 100

# Create the farm:
farm <- data.frame(id =1:n_cows,
                   age=round(runif(n_cows,730,1642)),
                   infected=0)

# We start by initially infected one cow - the first one:
farm$infected[1] <- 1

# We set the probability of infection:
ProbInfection <- 0.01

# We want to simulate 5 years:
end_time <- 5 * 365

# Collect the mean age of the cows in herd over time:
age_collect <- numeric(end_time)

# Collect the number of susceptible and infected animals at each time point:
inf_collect <- data.frame(susceptible=rep(0,end_time), infected=rep(0, end_time))

for (t in seq_len(end_time))
{

    # Identify the cows that are not infected (susceptible):
    PotInf       <- which(farm$infected==0)
    # Draw randomly which of the cows that should be (newly) infected:
    NewInf       <- PotInf[rbinom(length(PotInf),1,prob=ProbInfection)==1]
    # If there are new infections, then we infect them:
    if(length(NewInf)>0) farm$infected[NewInf] <- 1

   # Add one day to the age of all the animals, for each simulated:
   farm$age <- farm$age + 1

   # Save the daily mean age of all cows:
   age_collect[t] <- mean(farm$age)
   
   # Save the number of susceptible and infected:
   sus <- length(farm$infected[farm$infected==0])
   inf <- length(farm$infected[farm$infected==1])
   inf_collect[t,1] <- sus
   inf_collect[t,2] <- inf
}

```

And we plot the course of infection dynamics:

```{r}

plot(inf_collect$susceptible, type="l", lwd=2, col=3, xlab="Time", ylab="No. individuals", ylim=c(0,n_cows))
lines(inf_collect$infected, type="l", lwd=2, col=2)
legend("right", text.col=c(3,2), legend=c("Susceptible", "Infected"), 
       bty="n")

```

The number of susceptible individuals decrease over time and the number of infected individuals increase. 



# 2. Exercises

Hint: Modify the above code to solve the exercises.

## A.

ProbInf is the equivalent of the beta parameter commonly used in simulation models. It is the probability that a new infection will occur.

Try to adjust the ProbInf up and down. Can you make the number of susceptible and infected cross each other right in the middle of the simulation? How does the stochastic element come into play here?

`r solution_start()`
### Solution
It is difficult to control the stochastic process. However, parameters can be tweaked so they on average cross each other in the middle of the simulation.
`r solution_end()`

## B.
Try to make the probability of infection dependent on the number of infected individuals, for example like: Beta * Number of infected / total number of individuals.

`r solution_start()`
### Solution

```{r}
set.seed(250)

n_cows <- 100

# Create the farm:
farm <- data.frame(id =1:n_cows,
                   age=round(runif(n_cows,730,1642)),
                   infected=0)

# We start by initially infected one cow - the first one:
farm$infected[1] <- 1

# We set the probability of infection:
Beta <- 0.01

# We want to simulate 5 years:
end_time <- 5 * 365

# Collect the mean age of the cows in herd over time:
age_collect <- numeric(end_time)

# Collect the number of susceptible and infected animals at each time point:
inf_collect <- data.frame(susceptible=rep(0,end_time), infected=rep(0, end_time))

for (t in seq_along(end_time))
{
    # We update the ProbInf within the model:
    ProbInfection <- Beta * sum(farm$infected==1) / n_cows

    # Identify the cows that are not infected (susceptible):
    PotInf       <- which(farm$infected==0)
    # Draw randomly which of the cows that should be (newly) infected:
    NewInf       <- PotInf[rbinom(length(PotInf),1,prob=ProbInfection)==1]
    # If there are new infections, then we infect them:
    if(length(NewInf)>0) farm$infected[NewInf] <- 1

   # Add one day to the age of all the animals, for each simulated:
   farm$age <- farm$age + 1

   # Save the daily mean age of all cows:
   age_collect[t] <- mean(farm$age)
   
   # Save the number of susceptible and infected:
   sus <- length(farm$infected[farm$infected==0])
   inf <- length(farm$infected[farm$infected==1])
   inf_collect[t,1] <- sus
   inf_collect[t,2] <- inf
}


plot(inf_collect$susceptible, type="l", lwd=2, col=3, xlab="Time", ylab="No. individuals", ylim=c(0,n_cows))
lines(inf_collect$infected, type="l", lwd=2, col=2)
legend("right", text.col=c(3,2), legend=c("Susceptible", "Infected"), 
       bty="n")

```

`r solution_end()`


## C.
Modify the model so that there is a third compartment, Recovered (R) that infected individuals can enter with a certain probability.

`r solution_start()`
### Solution

```{r}
set.seed(250)

n_cows <- 100

# Create the farm:
farm <- data.frame(id =1:n_cows,
                   age=round(runif(n_cows,730,1642)),
                   infected=0)

# We now have three compartments in the "infected" column:
# 0 = Susceptible
# 1 = Infected
# 2 = Recovered

# We start by initially infected one cow - the first one:
farm$infected[1] <- 1

# We set the probability of infection:
Beta <- 0.01

# We set the probability of cure:
Gamma <- 0.03

# We want to simulate 5 years:
end_time <- 5 * 365

# Collect the mean age of the cows in herd over time:
age_collect <- numeric(end_time)

# Collect the number of susceptible and infected animals at each time point:
inf_collect <- data.frame(susceptible=rep(0,end_time), infected=rep(0, end_time), recovered=rep(0, end_time))

for (t in seq_len(end_time))
{
   # We update the ProbInf within the model:
   ProbInfection <- Beta * sum(farm$infected==1) / n_cows
   
   ### Infection ###
   
   # Identify the cows that are not infected (susceptible):
   PotInf       <- which(farm$infected==0)
   # Draw randomly which of the cows that should be (newly) infected:
   NewInf       <- PotInf[rbinom(length(PotInf),1,prob=ProbInfection)==1]
   # If there are new infections, then we infect them:
   if(length(NewInf)>0) farm$infected[NewInf] <- 1
   
   ### Recovery ###
   
   # Identify the cows that are infected:
   PotRec       <- which(farm$infected==1)
   # Draw randomly which of the cows that should recover:
   NewRec       <- PotRec[rbinom(length(PotRec),1,prob=Gamma)==1]
   # If there are any recoveries, they recover:
   if(length(NewRec)>0) farm$infected[NewRec] <- 2

   # Add one day to the age of all the animals, for each simulated:
   farm$age <- farm$age + 1
   
   # Save the daily mean age of all cows:
   age_collect[t] <- mean(farm$age)
   
   # Save the number of susceptible, infected and recovered:
   sus <- length(farm$infected[farm$infected==0])
   inf <- length(farm$infected[farm$infected==1])
   rec <- length(farm$infected[farm$infected==2])
   
   inf_collect[t,1] <- sus
   inf_collect[t,2] <- inf
   inf_collect[t,3] <- rec
}


plot(inf_collect$susceptible, type="l", lwd=2, col=3, xlab="Time", ylab="No. individuals", ylim=c(0,n_cows))
lines(inf_collect$infected, type="l", lwd=2, col=2)
lines(inf_collect$recovered, type="l", lwd=2, col=4)
legend("right", text.col=c(3,2, 4), legend=c("Susceptible", "Infected", "Recovered"), 
       bty="n")

```

`r solution_end()`

## D.
A more realistic scenario is that the individuals recover after a fixed number of days, for example 10 days. Try to include that in the model. Hint: You will need to set a counter to keep track of how many days the infected individuals have before they can recover. Plot the output.

`r solution_start()`
### Solution

```{r}
set.seed(250)

n_cows <- 100

# Create the farm:
farm <- data.frame(id =1:n_cows,
                   age=round(runif(n_cows,730,1642)),
                   infected=0, infected_days=0)

# We now have three compartments in the "infected" column:
# 0 = Susceptible
# 1 = Infected
# 2 = Recovered

# We start by initially infected one cow - the first one:
farm$infected[1] <- 1

# We set the probability of infection:
Beta <- 0.01

# We want to simulate 5 years:
end_time <- 5 * 365

# Collect the mean age of the cows in herd over time:
age_collect <- numeric(end_time)

# Collect the number of susceptible and infected animals at each time point:
inf_collect <- data.frame(susceptible=rep(0,end_time), infected=rep(0, end_time), recovered=rep(0, end_time))

for (t in seq_len(end_time))
{
   # We update the ProbInf within the model:
   ProbInfection <- Beta * sum(farm$infected==1) / n_cows
   
   ### Infection ###
   
   # Identify the cows that are not infected (susceptible):
   PotInf       <- which(farm$infected==0)
   # Draw randomly which of the cows that should be (newly) infected:
   NewInf       <- PotInf[rbinom(length(PotInf),1,prob=ProbInfection)==1]
   # If there are new infections, then we infect them:
   if(length(NewInf)>0) farm$infected[NewInf] <- 1
   
   # Identify the cows that have been infected for 10 days:
   Rec          <- which(farm$infected_days==1)
   # If there are any recoveries, they recover:
   if(length(Rec)>0) farm$infected[Rec] <- 2
   # And we set their infected counter to zero again:
   if(length(Rec)>0) farm$infetced_days[Rec] <- 0

   # And update the infected days counter (at each time step):
   farm$infected_days[farm$infected==1] <- farm$infected_days[farm$infected==1] + 1

   
   ### Recovery ###

   # Add one day to the age of all the animals, for each simulated:
   farm$age <- farm$age + 1
   
   # Save the daily mean age of all cows:
   age_collect[t] <- mean(farm$age)
   
   # Save the number of susceptible, infected and recovered:
   sus <- length(farm$infected[farm$infected==0])
   inf <- length(farm$infected[farm$infected==1])
   rec <- length(farm$infected[farm$infected==2])

   inf_collect[t,1] <- sus
   inf_collect[t,2] <- inf
   inf_collect[t,3] <- rec
}


plot(inf_collect$susceptible, type="l", lwd=2, col=3, xlab="Time", ylab="No. individuals", ylim=c(0,n_cows))
lines(inf_collect$infected, type="l", lwd=2, col=2)
lines(inf_collect$recovered, type="l", lwd=2, col=4)
legend("right", text.col=c(3,2, 4), legend=c("Susceptible", "Infected", "Recovered"), 
       bty="n")

```

`r solution_end()`


***

# 3. Bonus exercise

Modify the R6 model (bonus exercise from exercise 1) to include these infection dynamics. Include the Beta and Gamma parameters as arguments to the initialize method. Create a new method for infecting an animal, and another method for extracting the current total number of susceptible, infected and recovered animals.

`r solution_start()`
### Solution

We can simply extend the model we created for the previous session:

```{r}
library("R6")

## Define the farm class:
Farm <- R6Class("Farm",

  # Private objects that we can only see within the class:
  private = list(
    
    # The number of cows:
    n_cows = 0L,
    
    # The age of replacement heifers:
    age_heifer = 0L,
    
    # The age at which animals are replaced:
    age_replacement = 0L,
    
    # Beta:
    beta = 0.0,
    
    # Number of infected days:
    max_infected_days = 0L,
    
    # The data frame of cows:
    cows = data.frame(id = numeric(0),
                      age = numeric(0),
                      infected = numeric(0),
                      infected_days = numeric(0)),
    
    # The current time:
    time = 0L
    
  ),
                              
  # The public methods:
	public = list(

	  # An initialize method is mandatory:
		initialize = function(n_cows, age_heifer = 730, age_replacement = 1642, beta = 0.01, infected_days=10L) {
		  
		  # Set the number of cows:
		  private$n_cows <- n_cows
		  
		  # Store the ages using the public method we provide below:
		  self$set_replacement_ages(age_heifer = age_heifer, age_replacement = age_replacement)
		  
		  # Store beta:
		  self$set_beta(beta)
		  
		  # Store infected_days:
		  self$set_infected_days(infected_days)
		                
		  # Set the data frame of cows:
		  private$cows <- data.frame(id = 1:n_cows, 
		                  age = round(runif(n_cows,age_heifer,age_replacement)),
                      infected = 0,
                      infected_days = 0)
		  
		  # Set the time:
		  private$time <- 1L
		  
		},

		# A set method for the replacement ages:
		set_replacement_ages = function(age_heifer = NULL, age_replacement = NULL) {
		  if(!is.null(age_heifer)) private$age_heifer <- age_heifer
		  if(!is.null(age_replacement)) private$age_replacement <- age_replacement
		},

		# A set method for beta:
		set_beta = function(beta) {
		  private$beta <- beta
		},
		
		# A set method for infected_days:
		set_infected_days = function(infected_days) {
		  private$max_infected_days <- infected_days
		},
		
		# An infect method to change the status of a single cow:
		infect = function(cow) {
		  
		  private$cows$infected[cow] <- 1L
		  private$cows$infected_days[cow] <- 0L
		  
		},
		
		# An update method for a single time step:
		update = function() {
		  
		  ## First some sanity checks to help find bugs:
		  stopifnot(all(private$cows$infected %in% c(0,1,2)))
		  stopifnot(all(private$cows$infected_days >= 0))

		  
		  ## Do the infection part:

		  # Update prob_inf partly using a method we define below:
		  ProbInfection <- private$beta * self$get_totals()["infected"] / private$n_cows
		  
      # Identify the cows that are not infected (susceptible):
      PotInf <- which(private$cows$infected==0)
      # Draw randomly which of the cows that should be (newly) infected:
      NewInf <- PotInf[rbinom(length(PotInf),1,prob=ProbInfection)==1]
      # If there are new infections, then we infect them:
      if(length(NewInf)>0){
        private$cows$infected[NewInf] <- 1
        private$cows$infected_days[NewInf] <- 0
      }
  	  
		  ## Do the recovery part:
		  
      # Identify the cows that have been infected for 10 days:
      Rec <- which(private$cows$infected_days >= private$max_infected_days)
      # If there are any recoveries, they recover and we set their infected counter to zero again:
      if(length(Rec)>0) {
        private$cows$infected[Rec] <- 2
        private$cows$infetced_days[Rec] <- 0
      }

		  ## Then update counters:
		  
      # Update the infected days counter:
      private$cows$infected_days[private$cows$infected==1] <- private$cows$infected_days[private$cows$infected==1] + 1

		  # If cows reach the age of age_replacement, they are replaced with a new cow that is age_heifer days old:
      private$cows$age[private$cows$age>=private$age_replacement] <- private$age_heifer

		  private$cows$age <- private$cows$age + 1L
  	  private$time <- private$time + 1L

		},

		# A method to get the current cows:
		get_cows = function() {
		  
		  return(private$cows)
		  
		},
		
		# A method to get the current average age:
		get_mean_age = function() {
		  
		  mean_age <- mean(private$cows$age)
		  return(mean_age)
		  
		},
		
		# A method to get the number of susceptible, infected and recovered:
		get_totals = function() {
		  
		  stopifnot(all(private$cows$infected %in% c(0,1,2)))
		  rv <- c(
		    susceptible = sum(private$cows$infected==0L),
		    infected = sum(private$cows$infected==1L),
		    recovered = sum(private$cows$infected==2L)
		  )
		  
		  return(rv)
		  
		}

	)
)
```

The code to run this is very similar to before, except we now need to initialise infection and also capture the total susceptible/infected/recovered at each time step:

```{r}
# Create a new object, specifying beta explicitly so that it is more interesting:
farm <- Farm$new(n_cows = 100L, beta = 0.15)

end_time <- 5 * 365
age_collect <- numeric(end_time)
inf_collect <- data.frame(susceptible=rep(0,end_time), infected=rep(0, end_time), recovered=rep(0, end_time))

# First we must initialise infection in animal 1:
farm$infect(1L)

# Then we can loop over the other time points:
for (t in seq_len(end_time)) {
  
   # Update the object:
   farm$update()

   # Save the daily mean age of all cows:
   age_collect[t] <- farm$get_mean_age()

   # Save the infection statuses:
   inf_collect[t,1:3] <- farm$get_totals()
}

plot(inf_collect$susceptible, type="l", lwd=2, col=3, xlab="Time", ylab="No. individuals", ylim=c(0,n_cows))
lines(inf_collect$infected, type="l", lwd=2, col=2)
lines(inf_collect$recovered, type="l", lwd=2, col=4)
legend("right", text.col=c(3,2, 4), 
       legend=c("Susceptible", "Infected", "Recovered"), bty="n")
```

`r solution_end()`
